---
title: 'STA 602 Lab 6'
author: "Yicheng Shen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
---

```{r setup, message=F, warning=F, echo=F}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(rstanarm)
library(magrittr)
library(rstan)
library(bayesplot)
library(loo)
library(readxl)
library(coda)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(out.width = "80%", fig.align = 'center')
```

***

$$\begin{bmatrix} X \\ Y \\ Z \end{bmatrix} \sim N (\begin{bmatrix} 0 \\ 0 \\ 0  \end{bmatrix}, \begin{bmatrix} 1 & 0.9 & 0.1 \\  0.9 & 1 & 0.1 \\  0.1 & 0.1 & 1  \end{bmatrix} )$$

## Exercise 1

$$
\begin{aligned}
\bf U & = \begin{bmatrix} U_1 \\ U_2 \end{bmatrix} \sim N (\begin{bmatrix} \mu_1 \\ \mu_2 \end{bmatrix}, \begin{bmatrix} \Sigma_{11} & \Sigma_{12}\\ \Sigma_{21} & \Sigma_{22} \end{bmatrix} ) \\
U_1 | U_2 &= N(\mu, \Sigma) \\
\mu &= \mu_1 + \Sigma_{12} \Sigma^{-1}_{22}(U_2 - \mu_2) \\
\Sigma &= \Sigma_{11} - \Sigma_{12}\Sigma_{22}^{-1} \Sigma_{21}
\end{aligned}
$$

## Exercise 2

The three conditional distributions are specified below in the Gibbs sampler (using the formula above). 

## Exercise 3

The mixing for x and y draws are terrible. Their effective sample sizes are low as expected. We see very high, slowly decaying ACF as well. 

```{r, fig.height=6}
# initialization
x=y=z=0
# result
S = 1000
POST = matrix(NA, S, 3)
# gibbs sampling
for(i in 1:S){
# update x
x = rnorm(1,
          mean = 0 + matrix(c(0.9, 0.1), 1, 2) %*% solve(matrix(c(1, 0.1, 0.1, 1), 2, 2)) %*% matrix(c(y, z)) ,
          sd = sqrt(1 - matrix(c(0.9, 0.1), 1, 2) %*% solve(matrix(c(1, 0.1, 0.1, 1), 2, 2)) %*% matrix(c(0.9, 0.1)))
          )
# update y
y = rnorm(1,
          mean = 0 + matrix(c(0.9, 0.1), 1, 2) %*% solve(matrix(c(1, 0.1, 0.1, 1), 2, 2)) %*% matrix(c(x, z)) ,
          sd = sqrt(1 - matrix(c(0.9, 0.1), 1, 2) %*% solve(matrix(c(1, 0.1, 0.1, 1), 2, 2)) %*% matrix(c(0.9, 0.1)))
          )
# update z
z = rnorm(1,
          mean = 0 + matrix(c(0.1, 0.1), 1, 2) %*% solve(matrix(c(1, 0.9, 0.9, 1), 2, 2)) %*% matrix(c(x, y)) ,
          sd = sqrt(1 - matrix(c(0.1, 0.1), 1, 2) %*% solve(matrix(c(1, 0.9, 0.9, 1), 2, 2)) %*% matrix(c(0.1, 0.1)))
          )
# save
  POST[i,] = c(x, y, z)
}

colnames(POST) = c("x", "y", "z") 
coda_obj = mcmc(POST)
par(mfrow = c(3,1))
# trace plot
traceplot(coda_obj)
effectiveSize(coda_obj)
```


```{r}
# autocorrelation plots
autocorr.plot(coda_obj)
```


## Exercise 4

The blocked conditionals are shown below. 

## Exercise 5

The mixing is pretty good in this case. 

```{r, fig.height=6}
# initialization
x=y=z=0
# result
S = 1000
POST = matrix(NA, S, 3)
# gibbs sampling
for(i in 1:S){
# update x y 
  xy = mvtnorm::rmvnorm(1,
    c(0, 0) + matrix(c(0.1, 0.1), 2, 1) %*% 1 %*% matrix(z) ,
    (matrix(c(1, 0.9, 0.9, 1), 2, 2) - matrix(c(0.1, 0.1), 2, 1) %*% 1 %*% matrix(c(0.1, 0.1), 1, 2))
  )
  # update z
  z = rnorm(1,
    mean = 0 + 
      matrix(c(0.1, 0.1), 1, 2) %*% solve(matrix(c(1, 0.9, 0.9, 1), 2, 2)) %*% matrix(c(x, y)) ,
    sd = sqrt(1 - 
                matrix(c(0.1, 0.1), 1, 2) %*% solve(matrix(c(1, 0.9, 0.9, 1), 2, 2)) %*% matrix(c(0.1, 0.1), 2, 1))
  )

# save
  POST[i,] = c(xy, z)
}

colnames(POST) = c("x", "y", "z") 
coda_obj = mcmc(POST)
par(mfrow = c(3,1))
# trace plot
traceplot(coda_obj)
effectiveSize(coda_obj)
```


## Exercise 6

The first is not efficient due to the high dependence between x and y. While in 2nd case, (x,y) and z are less dependent on each other. 
