---
title: 'STA 602 Lab 2'
author: "Yicheng Shen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
---

```{r setup, message=F, warning=F, echo=F}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(magrittr)
library(plyr)
library(tufte)
library(knitr)
library(gridExtra)
library(rstanarm)
library(rstan)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(fig.align = 'center')
```

***

## Exercise 1

```{r}
tumors <- read.csv(file = url("http://www.stat.columbia.edu/~gelman/book/data/rats.asc"),
                     skip = 2, header = T, sep = " ")[,c(1,2)]
y <- tumors$y
N <- tumors$N
n <- length(y)
```

```{r, eval = F}
plot(seq(0, 1, length.out = 1000), 
     dbeta(seq(0, 1, length.out = 1000), 1, 1), type = 'l', 
     xlab = expression(theta), ylab = "Density",
     main = "The Beta(1, 1) density")
```

```{r, results = "hide"}
stan_dat <- list(n = n, N = N, y =y, a = 1, b = 1)
fit_pool <- stan('lab-02-pool.stan',data = stan_dat,chains = 2,refresh = 0)
pool_output <- rstan::extract(fit_pool)
```

```{r, out.width="70%", fig.align = "center"}
mean(pool_output$theta)
hist(pool_output$theta)
```

The distribution of $\theta$ seems approximately normal and centers around the mean of 0.155.


## Exercise 2

```{r, results = "hide"}
stan_dat <- list(n = n, N = N, y =y, a = 1, b = 1)
fit_nopool <- stan('lab-02-nopool.stan',data = stan_dat,chains = 2,refresh = 0)
nopool_output <- rstan::extract(fit_nopool)
```

```{r, out.width="70%", fig.align = "center"}
apply(nopool_output$theta, 2, mean)
boxplot(nopool_output$theta)
```

We have 71 groups, therefore for each group we have a $\theta_i$. The visualization above shows posterior distribution of each of the 71 $\theta_i$. Each point is a point from the posterior density of $\theta_i$.


## Exercise 3

pool stan code: 

```{r, eval = F}
parameters { 
  real<lower=0, upper=1> theta;  // chance of success (pooled) 
} 
```

nopool stan code: 

```{r, eval = F}
parameters { 
  vector<lower=0, upper=1>[n] theta;  // chance of success (unpooled) 
} 
```

The pool model uses `real` while the no pool code uses `vector` and `[n]`. The difference affects whether we view the groups behave independently (each group has varying parameter) or they share the same parameter. 


## Exercise 4

```{r, out.width="70%", fig.align = "center"}
par(mfrow = c(4, 4))
par(mar=c(2,2,2,2))
for(a_val in c(1, 10, 25, 100)){
  for(b_val in rev(c(1, 10, 25, 100))){
    plot(seq(0, 1, length.out = 1000), 
     dbeta(seq(0, 1, length.out = 1000), a_val, b_val),
     type = 'l', 
     xlab = expression(theta), ylab = "Density",
     main = paste0("Beta(", a_val, ", ", b_val, ")"))
  }
}
```


```{r}
output_list <- list()
for(a_val in c(1, 10, 25, 100)){
  for(b_val in c(1, 10, 25, 100)){
    stan_dat <- list(n = n, N = N, y = y, a = a_val, b = b_val)
    fit_pool <- stan('lab-02-pool.stan', data = stan_dat, chains = 2, refresh = 0)
    output_list[[paste0("a_", a_val, ":b_", b_val)]] <- rstan::extract(fit_pool)[["theta"]]
  }
}
```


```{r, out.width="70%", fig.align = "center"}
output_list %>%
  plyr::ldply(function(theta){
    reshape2::melt(theta) %>%
      dplyr::mutate(post_mean = mean(theta))
  }, .id = "prior") %>%
  tidyr::separate("prior", into = c("a", "b"), sep = ":") %>%
  dplyr::mutate(a = as.numeric(gsub("._", "", a)),
                b = as.numeric(gsub("._", "", b))) %>%
  ggplot2::ggplot() +
  geom_density(aes(x = value)) +
  geom_vline(aes(xintercept = post_mean)) +
  facet_grid(a~factor(b, levels = rev(c(1, 10, 25, 100)))) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = expression(theta), y = "Density")
```


a represents the number of successes in prior observations, b represents the number of failures in the prior observations. 

## Exercise 5

```{r, out.width="70%", fig.align = "center"}
a <- ggplot(tumors) + geom_point(aes(x = N, y = y))
b <- ggplot(tumors) + geom_point(aes(x = N, y = y / N))
grid.arrange(a, b, nrow = 1)
tumors %>% summarise(proportion = mean(y/N))
```

In the `tumors` data, we observe a relatively low proportion of successes compared with failures. The overall average $\frac{y}{N}$ is 0.1381151. 


## Exercise 6

Most prior beliefs (represented by a and b) are not very close to the data, but a Beta(1,10) or Beta(10,100) prior may be a close match to the data. 


## Exercise 7

```{r}
# approach 1
mle.1 <- sum(y)/sum(N)
mle.1
mean(pool_output$theta)

# approach 2
mle.2 <- y/N
mle.2
apply(nopool_output$theta, 2, mean)
```

As mentioned above, the posterior mean is a weighted average of the prior mean and the MLE. When our prior mean is determined by Beta(1,1), the posterior from Bayesian approach is close to the MLE. 

***