---
title: 'STA 602 Lab 1'
author: "Yicheng Shen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
---

```{r setup, message=F, warning=F, echo=F}
library(tidyverse)
require(magrittr)
require(plyr)
library(tufte)
library(knitr)
library(gridExtra)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(fig.align = 'center')
```

Here's a list of resources for learning rmarkdown:

1. [R markdown introduction](https://rmarkdown.rstudio.com/lesson-1.html)
1. [R for Data Science](https://r4ds.had.co.nz/)
1. [Introduction to R markdown](https://rmarkdown.rstudio.com/articles_intro.html)
1. [Another introduction to R markdown](https://rpubs.com/andrew-cho/303981)
1. [R markdown cheatshett](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
1. [Data visualization with ggplot2 cheat sheet](https://rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)
1. [Learn LaTex in 30 minutes](https://www.overleaf.com/learn/latex/Learn_LaTeX_in_30_minutes)


***

## Exercise 1

1. Create a code chunk and set the header parameter to TRUE and print out the top rows of the table with `head()` as above.

```{r load_data_with_header, eval = T}
data <- read.table(file = url("http://www2.stat.duke.edu/~pdh10/FCBS/Exercises/azdiabetes.dat"), 
                   header = TRUE)
head(data)
```


## Exercise 2

2. Generate a sequence of $100$ equispaced real numbers from $0$ to $1$ and store it in a variable called `seq2`.

```{r seq2, eval = T}
seq2 <- seq(from = 0, to = 1, by = 1 / 99)  # If use 1/100 would give 101 numbers, so 1/99
length(seq2)
```


## Exercise 3

3. Sort the entries in `seq3` from greatest to least.

```{r sort seq3, eval = T}
seq3 <- seq(from = -3, to = 3, by = .5)
sort(seq3, decreasing = T)
```


## Exercise 4

4. Find the variance of each row of `mat5`

```{r var, eval = T}
mat5 <- matrix(seq(1, 100, 1), nrow = 4, ncol = 25, byrow = T)
apply(X = mat5, MARGIN = 1, FUN = var)
```


## Exercise 5 

5. Generate $500$ samples from a Beta distribution with shape parameter $[a, b] = [0.5, 0.5]$ and store the samples in a variable called `W`

```{r rbeta, eval = T}
set.seed(32507) 
W <- rbeta(500, 0.5, 0.5) 
head(W) # output is big hence not shown completely
```


## Exercise 6

6. Browse online resources (some below), or use code from above to make a few plots of your own.

```{r, out.width="90%", eval = T}
set.seed(253)
norm_samples <- rnorm(10000)

a <- norm_samples %>%
  data.frame(x = .) %>%
  mutate(`x^2` = x^2, `x^3` = x^3, `abs(x)` = abs(x)) %>%
  ggplot2::ggplot() + 
  geom_point(aes(x,`x^2`),color = "blue") + 
  geom_point(aes(x,`abs(x)`),color = "red") +
  geom_point(aes(x,`x^3`),color = "orange") + 
  labs(title = "Transformation of Normal Variable", y = "") + theme_bw()

b <- norm_samples %>%
  data.frame(x = .) %>%
  mutate(`Cumulative Sums` = cumsum(x), time = 1:10000) %>%
  ggplot2::ggplot() + geom_line(aes(time, `Cumulative Sums`)) +
  labs(title = "Random Time Series Plot") + theme_bw()

set.seed(466)
norm_samples_2 <- rnorm(10000)
poission_samples_2 <- rpois(10000, 10)

c <- data.frame(x = norm_samples, y = norm_samples_2, z = poission_samples_2) %>%
  ggplot2::ggplot() + geom_jitter(aes(x,y,color = z)) + 
  labs(title = "Random Scatterplot") + theme_bw()

d <- bayesrules::plot_beta_binomial(alpha = 20, beta = 20, y = 40, n = 100) +
  labs(title = "Beta Binomial Model") + theme_bw()

grid.arrange(a, b, c, d, nrow = 2)
```
***