---
title: 'STA 602 Lab 5'
author: "Yicheng Shen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
---

```{r setup, message=F, warning=F, echo=F}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(rstanarm)
library(magrittr)
library(rstan)
library(bayesplot)
library(loo)
library(readxl)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(out.width = "60%", fig.align = 'center')
```

***

## Exercise 1

```{r}
x <- runif(1000,-1,1)
y <- runif(1000,-1,1)
idx <- (x^2 + y^2 < 1)

plot(x[idx], y[idx], xlab="x", ylab="y")
curve((  1 * (1 - x^2)^0.5 ), add=TRUE, from=-1 , to =1)
curve(( -1 * (1 - x^2)^0.5 ), add=TRUE, from=-1 , to =1)
```

Can we use Exponential(1) as the candidate density to draw samples for N(0,1) on the interval (-3,3)? No. 
```{r}
S <- 5000
plot(density(rexp(S,1)))
lines(density(rnorm(S)))
```
Can we use N(0,1) as candidate density to draw samples from the standard Cauchy distribution on the interval (-2,2)? Yes.

```{r}
plot(density(rnorm(S)), ylim = c(0,1))
lines(density(rcauchy(S, -2, 2)))
```

If we only know the kernel of a uni-variate posterior distribution we are interested in, can we use RS? Yes. 

What if it’s a 100-variate posterior distribution? No. 


## Exercise 2

```{r}
theta <- seq(0,1,length.out = 200)
g_theta <- sin(pi * theta)^2

plot(theta, g_theta, type = "l", col="red",
     xlab = expression(theta), ylab = expression(sin(pi~theta)^2))
```

```{r}
# target density function
f <- function(x){
  sin(pi*x)^2
}

nsim <- 10000

# acceptance count, total count
ac <- tc <- 0

# to store results
res <- numeric(nsim)

# candidate density function
g <- function(x) { dunif(x) }

M <- 0.0001 

while (ac < nsim){
  tc <- tc + 1
  x <- runif(1)
  u <- runif(1)
  
  if ( u < f(x) )
  {
    res[ac] <- x
    ac <- ac + 1
  }

}
```

```{r}
plot(theta, g_theta*2, type = "l", col="red",
     xlab = expression(theta), ylab = expression(sin(pi~theta)^2), ylim = c(0,2))
lines(density(res))
```


## Exercise 3

one is the density of distribution one is kernel. 

## Exercise 4



## Exercise 5



## Exercise 6


the ratio of normalizing constant

$1/N \sum \frac{\tilde f}{\tilde g} h(x')$ converge to $E_f [h]$


## Exercise 7


## Exercise 8


Plug in posterior distributions.




## Exercise 9


```{r, eval=F}
tumors <- read.csv(file = url("http://www.stat.columbia.edu/~gelman/book/data/rats.asc"),
                     skip = 2, header = T, sep = " ")[,c(1,2)]
y <- tumors$y
N <- tumors$N
n <- length(y)

stan_dat <- list(n = n, N = N, y =y, a = 1, b = 1)
fit_pool <- stan('lab-02-pool.stan', data = stan_dat, chains = 2, refresh = 0)
## Trying to compile a simple C file
pool_output <- rstan::extract(fit_pool)
```


```{r, eval=F}
as <- c(1,10,25,100)
bs <- c(1,10,25,100)

# posterior parameters under the current Beta(1,1) prior
post_a0 <- 1 + sum(y)
post_b0 <- 1 + sum(N) - sum(y)

IS_sensi_mean <- matrix(0, length(as), length(bs))  # IS estimated posterior mean
exp_sensi_mean <- matrix(0, length(as), length(bs))  # theoretical posterior mean

for (i in (1:length(as))){
  for (j in (1:length(bs))){

    # fill in your codes here
    
  }
}

# plot the theoretical posterior means against the IS estimated posterior means
plot(c(IS_sensi_mean), c(exp_sensi_mean), xlab = "IS", ylab = "Expected")
abline(a = 0, b = 1)  # 45 degree line
```



```{r}
x=40; n=100
# p is the target distribution to sample from
p = function(theta) {
theta^x*(1-theta)^(n-x) *exp(-(theta-0.5)^2)
}
# q is something easy to sample
q = function(x) { dunif(x) }
# Choose a constant that satisfies f<M*g,
# but make M as small as possible
# In finding this value I ``cheated''
M=7e-30
```

```{r}
S=10000
# draw from q
theta.q = runif(S)
# compute acceptance probability
acc.prob = p(theta.q)/(M*q(theta.q))
# indicator for acceptance
acc.ind = rbinom(S,size=1,prob=acc.prob)
# proportion of accepted draws
mean(acc.ind)
## [1] 0.1018
# the accepted draws
theta.p = theta.q[as.logical(acc.ind)]
```

